Rényi Entropy-Based Heterogeneity Detection in SAR Data

### 2025-03-15 Text/Articles/RENYI_ENTROPY_SAR_2025/R0-Renyi-entropy-sar.qmd
Started revision

The main distributions considered for SAR intensity data are the $\Gamma_{\text{SAR}}$ distribution, which is suitable for fully developed speckle, and the $\mathcal{G}^0_I$ distribution, which is able to describe roughness\ [@Frery1997]. These distributions are characterized by the following probability density functions (pdfs):
\begin{align}
	f_{\Gamma_{\text{SAR}}}\bigl(z;L, \mu \bigr) 
    &= \frac{L^L}{\Gamma(L)\,\mu^L} z^{L-1} 
    \exp \biggl(-\frac{Lz}{\mu}\biggr)
    \mathbbm 1_{\mathbbm R_+}(z) \label{E:gamma1}
    \intertext{and}
    f_{\mathcal{G}^0_I}\bigl(z; \mu, \alpha, L \bigr) 
    &= \frac{L^L\,\Gamma(L-\alpha)}
    {\bigl[-\mu(\alpha+1)\bigr]^{\alpha} \Gamma(-\alpha)\,\Gamma(L)}
    \notag \\
    &\quad \times
    \frac{z^{L-1}}
    {\bigl[-\mu(\alpha+1)+Lz\bigr]^{L-\alpha}}
    \mathbbm 1_{\mathbbm R_+}(z), \label{E:gi01}
\end{align}
where $\mu > 0$ is the mean,
$\alpha < 0$ measures the roughness, $L \geq 1$ is the number of
looks, $\Gamma(\cdot)$ is the gamma function, and
$\mathbbm 1_{A}(z)$ is the indicator function of the set $A$. 
As demonstrated by\ [@Frery1997], the $\Gamma_{\text{SAR}}$  model is a particular case of the $\mathcal{G}^0_I$ distribution. Specifically, for a given $\mu$ fixed,
$$
f_{\mathcal{G}^0_I}\big(z; \mu, \alpha, L\big)
\longrightarrow 
f_{\Gamma_{\text{SAR}}}(z;L, \mu) 
$$
when $\alpha\to-\infty$.

@@

Assume we have an estimator $\widetilde{H}_{\lambda}$ for the Rényi entropy of an arbitrary model. For testing \eqref{eq:hypothesis_test}, this estimator is expected to be close to the theoretical expression given in Equation \eqref{eq-HGammaSAR}. 
Since $L\geq1$ is known, we define the test statistic as follows:
\begin{multline}
\label{eq-test}
S(\bm{Z}; L) = \widetilde{H}_{\lambda} - \bigl\{\ln \widehat{\mu} - \ln L + \frac{1}{1-\lambda}
\bigl[-\lambda\,\ln\Gamma(L) \\  
+ \ln\Gamma\bigl(\lambda(L-1)+1\bigr)  
- \bigl(\lambda(L-1)+1\bigr)\,\ln\lambda
\bigr]\bigr\}.
\end{multline}
where $\ln \widehat{\mu}={\textstyle\frac{1}{n}\sum_{i=1}^n Z_{i}}$ is the natural logarithm of the sample mean.

When the null hypothesis holds, this test statistic is expected to be close to zero.
For illustrative purposes, @fig-density_entropyR shows the empirical density of $S(\bm{Z}; L)$, based on $10,000$ simulations of the $\Gamma_{\text{SAR}}$ model, with $\lambda=0.9$ and $L \in \{5,18\}$.
```{r Simulated_densityR, echo=FALSE, message=FALSE}
set.seed(1234567890, kind = "Mersenne-Twister")

@@

```
Under the asymptotic properties of the entropy estimator\ [@vasicek1976test], for sufficiently large samples, $S(\bm{Z}; L)$ follows an asymptotic normal distribution:
\begin{equation*}
S(\bm{Z}; L) \overset{d}{\longrightarrow} \mathcal{N}(\mu_S,\,\sigma^{2})\,.
\end{equation*}
We define the standardized test statistic:
\begin{equation*}
\varepsilon = \frac{S(\bm{Z}; L) - \hat{\mu}_S}{\hat{\sigma}_S},
\end{equation*}
where $\hat{\mu}_S = \mathbb{E}[S(\bm{Z}; L)]$ and $\hat{\sigma}_S = \sqrt{\text{Var}[S(\bm{Z}; L)]}$.
By the central limit theorem, under the null hypothesis, $\varepsilon$ asymptotically follows a standard normal distribution:
\begin{equation*}
\varepsilon \overset{d}{\longrightarrow} \mathcal{N}(0,1)\,,
\end{equation*}
where $d$ represents convergence in distribution.
@fig-density_entropyR presents the behavior of the empirical distribution of the standardized test statistic.
Thus, the $p$-values are calculated as $2\Phi(-|\varepsilon|)$,
where $\Phi(\cdot)$ is the cumulative distribution function of the standard normal distribution.
```{r Simulated_densityR_standardized0, echo=FALSE, message=FALSE}
set.seed(1234567890, kind = "Mersenne-Twister")

@@

Let $Z \sim \mathcal{G}^0_I(\alpha, \gamma, L)$, defined in\ [@Frery2024],  we compute $I = \int_0^\infty [f_{\mathcal{G}^0_I}(z)]^\lambda dz$. Using the pdf  
\begin{equation*}
f_{\mathcal{G}^0_I}(z) = C \frac{z^{L-1}}{(\gamma + Lz)^{L-\alpha}},  
\quad C = \frac{L^L \Gamma(L - \alpha)}{\gamma^\alpha \Gamma(-\alpha) \Gamma(L)}, \label{eq:pdf_G0I}
\end{equation*}
and setting $t = Lz/\gamma$, we obtain  
\begin{equation*}
I = C^\lambda \frac{\gamma^{1-\lambda+\lambda\alpha}}{L^{\lambda L + 1 - \lambda}}  
\int_0^\infty \frac{t^{\lambda(L-1)}}{(1+t)^{\lambda(L-\alpha)}} dt.
\end{equation*}
Applying the Beta function identity, $B(a,b) = \int_0^\infty \frac{t^{a-1}}{(1+t)^{a+b}} dt$ with $a = \lambda(L-1) + 1$, $b = \lambda(-\alpha+1) - 1$, we get  
\begin{equation*}
I= \gamma^{\,1 - \lambda}\,
   L^{\,\lambda - 1}
   \Bigl(\tfrac{\Gamma(L - \alpha)}{\Gamma(-\alpha)\,\Gamma(L)}\Bigr)^\lambda
   \,B(a,b).
\end{equation*}
Using the Rényi entropy definition in \eqref{E:entropy2}: $H_\lambda(Z) = \frac{1}{1-\lambda} \ln I$,
and substituting $\gamma = -\mu(\alpha + 1)$, we obtain the final expression for $\mathcal{G}^0_I(\mu, \alpha, L)$ in \eqref{eq-HGI0}.

Code is available at: <https://github.com/rjaneth/renyi-entropy>

>>>>

The main distributions considered for SAR intensity data are the $\Gamma_{\text{SAR}}$ distribution, which is suitable for fully developed speckle, and the $\mathcal{G}^0_I$ distribution, which is able to describe roughness\ [@Frery1997]. These distributions are characterized by the following probability density functions (pdfs):
\begin{equation}
	f_{\Gamma_{\text{SAR}}}\bigl(z;L, \mu \bigr) 
    = \frac{L^L}{\Gamma(L)\,\mu^L} z^{L-1} 
    \exp \biggl(-\frac{Lz}{\mu}\biggr)
    \mathbbm 1_{\mathbbm R_+}(z), \label{E:gamma1}
\end{equation}
and
\begin{multline}
    f_{\mathcal{G}^0_I}\bigl(z; \mu, \alpha, L \bigr) 
    = \frac{L^L\,\Gamma(L-\alpha)}
    {\bigl[-\mu(\alpha+1)\bigr]^{\alpha} \Gamma(-\alpha)\,\Gamma(L)}\\
    \frac{z^{L-1}}
    {\bigl[-\mu(\alpha+1)+Lz\bigr]^{L-\alpha}}
    \mathbbm 1_{\mathbbm R_+}(z), \label{E:gi01}
\end{multline}
where $\mu > 0$ is the mean,
$\alpha < 0$ measures the roughness, $L \geq 1$ is the number of
looks, $\Gamma(\cdot)$ is the gamma function, and
$\mathbbm 1_{A}(z)$ is the indicator function of the set $A$. 
As demonstrated by\ [@Frery1997], the $\Gamma_{\text{SAR}}$  model is a particular case of the $\mathcal{G}^0_I$ distribution. Specifically, for a given $\mu$ fixed,
$$
f_{\mathcal{G}^0_I}\big(z; \mu, \alpha, L\big)
\longrightarrow 
f_{\Gamma_{\text{SAR}}}(z;L, \mu) 
$$
when $\alpha\to-\infty$.

@@

Assume we have an estimator $\widetilde{H}_{\lambda}$ for the Rényi entropy of an arbitrary model. For testing \eqref{eq:hypothesis_test}, this estimator is expected to be close to the theoretical expression given in Equation \eqref{eq-HGammaSAR}. 
Since $L\geq1$ is known, we define the test statistic as follows:
\begin{multline}
\label{eq-test}
S(\bm{Z}; L) = \widetilde{H}_{\lambda} - \bigl\{\ln \widehat{\mu} - \ln L + \frac{1}{1-\lambda}
\bigl[-\lambda\,\ln\Gamma(L) \\  
+ \ln\Gamma\bigl(\lambda(L-1)+1\bigr)  
- \bigl(\lambda(L-1)+1\bigr)\,\ln\lambda
\bigr]\bigr\},
\end{multline}
where $\widehat{\mu}={n}^{-1}\sum_{i=1}^n Z_{i}$ is the sample mean.

When the null hypothesis holds, this test statistic is expected to be close to zero.
@fig-density_entropyR shows the empirical density of $S(\bm{Z}; L)$, based on $10,000$ simulations of the $\Gamma_{\text{SAR}}$ model, with $\lambda=0.9$ and $L \in \{5,18\}$.
```{r Simulated_densityR, echo=FALSE, message=FALSE}
set.seed(1234567890, kind = "Mersenne-Twister")

@@

```
Under the asymptotic properties of the entropy estimator\ [@vasicek1976test], for sufficiently large samples, $S(\bm{Z}; L)$ follows an asymptotic normal distribution:
\begin{equation*}
S(\bm{Z}; L) \overset{d}{\longrightarrow} \mathcal{N}(\mu_S,\,\sigma^{2})\,.
\end{equation*}
We define the standardized test statistic:
\begin{equation*}
\varepsilon = \frac{S(\bm{Z}; L) - \hat{\mu}_S}{\hat{\sigma}_S},
\end{equation*}
where $\hat{\mu}_S = \mathbb{E}[S(\bm{Z}; L)]$ and $\hat{\sigma}_S = \sqrt{\text{Var}[S(\bm{Z}; L)]}$.
By the central limit theorem, under the null hypothesis, $\varepsilon$ asymptotically follows a standard normal distribution:
\begin{equation*}
\varepsilon \overset{d}{\longrightarrow} \mathcal{N}(0,1)\,,
\end{equation*}
where $d$ represents convergence in distribution.
@fig-density_entropyR shows smoothed histograms of standardized test statistic.
Thus, the $p$-values are calculated as $2\Phi(-|\varepsilon|)$,
where $\Phi(\cdot)$ is the cumulative distribution function of the standard normal distribution.
```{r Simulated_densityR_standardized0, echo=FALSE, message=FALSE}
set.seed(1234567890, kind = "Mersenne-Twister")

@@

Let $Z \sim \mathcal{G}^0_I(\alpha, \gamma, L)$ as defined in Ref.\ [@Frery2024].
We compute $I = \int_0^\infty [f_{\mathcal{G}^0_I}(z)]^\lambda dz$ using the pdf  
\begin{equation*}
f_{\mathcal{G}^0_I}(z) = C \frac{z^{L-1}}{(\gamma + Lz)^{L-\alpha}},  
\quad C = \frac{L^L \Gamma(L - \alpha)}{\gamma^\alpha \Gamma(-\alpha) \Gamma(L)}. \label{eq:pdf_G0I}
\end{equation*}
Setting $t = Lz/\gamma$, we obtain  
\begin{equation*}
I = C^\lambda \frac{\gamma^{1-\lambda+\lambda\alpha}}{L^{\lambda L + 1 - \lambda}}  
\int_0^\infty \frac{t^{\lambda(L-1)}}{(1+t)^{\lambda(L-\alpha)}} dt.
\end{equation*}
Applying the Beta function identity, $B(a,b) = \int_0^\infty \frac{t^{a-1}}{(1+t)^{a+b}} dt$ with $a = \lambda(L-1) + 1$, $b = \lambda(-\alpha+1) - 1$, we get  
\begin{equation*}
I= \gamma^{\,1 - \lambda}\,
   L^{\,\lambda - 1}
   \Bigl(\tfrac{\Gamma(L - \alpha)}{\Gamma(-\alpha)\,\Gamma(L)}\Bigr)^\lambda
   \,B(a,b).
\end{equation*}
Using the Rényi entropy definition in \eqref{E:entropy2}: $H_\lambda(Z) = (1-\lambda)^{-1} \ln I$,
and substituting $\gamma = -\mu(\alpha + 1)$, we obtain the final expression for $\mathcal{G}^0_I(\mu, \alpha, L)$ in \eqref{eq-HGI0}.

The code is available at: <https://github.com/rjaneth/renyi-entropy>
